Title here

### 2020-09-23 analysis/paper/paper.Rmd
llegamos a la conclusion!

---
twocol: FALSE  #  Use FALSE for submission. 
title: Title here
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro Díaz
affiliation: "CIMA UBA blablabla"
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: AsymSAM.bib
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
---

```{r setup, echo = FALSE,warning=FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.path = "",
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.extra = 42,
  cache.path = "../cache/",
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  out.extra = ""
)
library(metR)
library(data.table)
library(ggplot2)
library(magrittr)
library(ggperiodic)
library(asymsam)
library(latex2exp)
library(tagger)

@@

@@

We defined the Southern Annular Mode (SAM) as the leading EOF of the monthly anomalies of geopotential field at 700 hPa south of 20\degree S (citation?). The EOF was performed by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude). The values where weighted by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999]. This same method was used at the rest of the levels considered in this paper.

@@

We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach2020.] for the period 1979 to 2018. Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0 degree latitude x 5.0 degree longitude global grid [@vose2012; @smith2008]. The same analysis was carried out using CRUTEM4 [@osborn2014] (not shown). We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude. CPCC: [schneider2015]  #FIXME

@@

Explain multiple regression interpretation #FIXME.

We adjusted p-values for False Detection Rate following @wilks2016.

@@

```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM and density estimates.", fig.width=6, fig.height=3, fig.env="figure*"}
indices %>% 
  .[lev %in% c(50, 700) & term != "full"] %>% 
  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
  ggplot(aes(time, estimate_norm)) +
  geom_line(aes(color = term)) +
  ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
                            aes(color = term, width = ..density.., x = min(time - 1e7))) +
  # geom_smooth(aes(color = term), method = "loess", n = 481, span = 10*12/481, method.args = list(degree = 1)) +
  # geom_point(data = ~.x[lev == 50][time == as.Date("1987-11-01") & term == "sym" | 
  #                         time == as.Date("1987-08-01") & term == "asym"]) +
  ggforce::geom_mark_circle(data = ~.x[lev == 50][time == as.Date("1987-11-01") & term == "sym" | 
                                                    time == as.Date("1987-08-01") & term == "asym"],
                            expand = unit(2, "mm")) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
  scale_y_continuous(NULL) +
  # scale_x_datetime(NULL, expand = c(0, 0)) +
  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
  tag_facets()
```

The temporal evolution of the Assymmetric and Symmetric SAM was firstly asssesed. Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 and 50hPa and their corresponding density estimates. Month-to-month variability is evident for both indices, with noisy variations in the low frequency. At first glance the series can be distinguished by their distributions. Whereas the tropospheric indices are aproximately normally distributed, the stratospheric indices are more long-tailed; that is, extreme values (both negative and positive) abound. The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.

The stratospheric Symmetric SAM varies strongly witha two-year period, which can be seen using spectral methods (Figure A3) or in the autocorrelation structure (Figure A4). There is a local peak at 2 years in the periodigram of the tropospheric Symmetric SAM also, although it's not statistically significant. In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.

@@

From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987 marked with a circle). We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis. At the same time, one-month-lag correlations increase with height. As a consequence, statospheric Assymmetric index tend to precede corresponding Symmetric index.

@@

Figure \ref{fig:cross-correlation}a) shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices. For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposfere. Above 100 hPa correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability. Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere. Examining panels b and c, we see that the Asymemtric and Symmetric SAM  share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour. Stratospheric coherency is stronger for the Asymmetric SAM than the Symemtric SAM. The stratospheric Symmetric SAM seems to connect more strongly to the trosposphere than the Asymmetric SAM.

@@

```{r trends, fig.cap = "Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
annual_trend <- indices %>%
  copy() %>% 
  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  .[, season := "Annual"]

indices %>% 
  copy() %>% 
  # .[time > as.Date("1990-01-01")] %>%
  .[, season := season(time)] %>% 
  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  rbind(annual_trend) %>% 
  .[, estimate := estimate*3600*24*365*10] %>% 
  .[, std.error := std.error*3600*24*365*10] %>% 
  setnames("type", "term") %>% 
  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
  ggplot(aes(lev, estimate)) +
  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) +
  geom_line() +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  geom_hline(yintercept = 0) +
  scale_x_level() +
  scale_y_continuous("Standard deviations per decade") +
  coord_flip() +
  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
  tag_facets("rc") 
```

The trends for each of the indices (Full, Symmetric, Assymetric) were evaluate for the whole period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters. As previously documented (e.g. @fogt2020), there is a statistically significant increase towards more positive SAM (panel a.1), which is evident only in Summer and Autumn (panels b.1 and c.1). We observe these increases mainly in the troposphere, reaching their maximum at at 100 hPa in Summer. By separating the SAM signal in its Asymmetric and Symmetric parts, we can not only see that these trends are almost entirely due to the Symmetric component (columns 2 vs. columns 3), but in some cases the trends become more clear. In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the signal; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel b.1 and b.3). In Autumn, using the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index.

@@

```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef. Points marked on panel b.2 are the location of the reference points used by \\cite{raphael04} for its Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
indices_2d_yearly %>% 
  .[lat <= -20] %>% 
  .[lev %in% c(50, 700)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, exclude =  0),
                      size = 0.3,
                      global.breaks = FALSE) +
  geom_qmap(~.x[lat <= -20]) +
  geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
                               lat = -49, term = "asym", lev = 700), 
             shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
                       limits = c(-45, 45),
                       oob = scales::squish,
                       breaks = AnchorBreaks(0, 10, 0)) +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr") +
  axis_labs_smol  +
  no_grid
```

@@

In the troposphere, panel a.2 shows the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3. The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures. The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel c.2). The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure. Thus, the tropospheric Asymmetric SAM index responds to zonal displacements in the position of the climatological wave-3 pattern, previously described in the literature (e.g. ....).

@@

The amplitude of first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean. This Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (column a) is very similar to the sum of the waves of the Asymmetric and Symmetric field (columns b and c, respectively). Column b of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b.1), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel b.2).

@@

To analyse the vertical structure of the geopotential anomalies asociated with the asymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65ªS and 40ªS for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}). The geopotential anomalies associated with the stratospheric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM. The vertical structure of this signal tilts about 60ª to the West between 100 hPa and 1 hPa, suggesting baroclinic processes and poleward transport of heat (#FIXME is this ok?). Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.

@@

The structures shown in panels a and b in  Figure \ref{fig:vertical-regression} are surprisignly robust to the choice of index level. For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a. Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b. Pattern cross-correlation is greater than 0.9 within stratospheric index levels and tropospheric index levels and less than .4 accross stratospheric and tropospheric levels (Figure A8), i.e. regressions obtained using Asymmetric SAM indices in the troposphere are very similar to each other and different to the ones obtaines using indices in the stratosphere and vice versa. The patterns mainly change in amplitude. The tropospheric pattern is maximised by the 300 hPa Asymmetric SAM index and the stratospheric pattern increased monotonically with height.

```{r enso-cor}
mei_sam_cor <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), mei = MEI)] %>% 
  .[indices, on = "time"] %>% 
  .[, with(cor.test(estimate, mei), 
           list(cor = estimate, 
                low = conf.int[1], 
                upp = conf.int[2])), by = .(lev, term)] 
# enso_sam_cor %>% 
#   ggplot(aes(lev, cor)) +
#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha = 0.2) +
#   geom_line(aes(color = term)) +
#   scale_x_level() +
#   coord_flip()

cor_texts <- mei_sam_cor[lev == 700, 
                          paste0(signif(cor, 2)), # " (CI: ", signif(low, 2), "–", signif(upp, 2), ")"), 
                          by = term] %>% 
  dcast(1~term)

>>>>

---
twocol: TRUE  #  Use FALSE for submission. 
title: Title here
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro Díaz
affiliation: "CIMA UBA blablabla"
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: AsymSAM.bib
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
---

```{r setup, echo = FALSE,warning=FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.path = "",
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.extra = 42,
  cache.path = "../cache/",
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  out.extra = ""
)

knitr::opts_hooks$set(fig.cap = function(options) {
  options$cache <- FALSE
  options
})

library(metR)
library(data.table)
library(ggplot2)
library(magrittr)
library(ggperiodic)
library(asymsam)
library(latex2exp)
library(tagger)
library(patchwork)
library(knitr)

@@

To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.

We describe the relationship between the SAM indices and temperature and precipitation. We use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree lagitude by 5\degree longitude. For precipitation, we use monthly, 0.5\degree latitude by 0.5\degree longitude data from the Global Precipitation Climatology Centre [@schneider2015].

@@

Traditionally the Souther Annular Mode (SAM) is defined as de leading Empiral Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels [@ho2012]. Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level. We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude). We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].

@@

Our method assumes linearity in the asymmetric component of the SAM. That is, we assume that zonal symmetries associated with positive SAM are oposite and equal to the ones associated with negatie SAM. @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites. Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (smaller than negative 1 standard deviation) show relatively high pattern correlations all seasons and are visually very linear (Figure A9). Therefore, we belive that our method is at the very least a reasonable approximation of the phenomenon.

By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons. Geopotential zonal anomalies computed by projecting the first EOF *of each season* are very similar to each other (Figure A10) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (between MAM and SON). Based on this, we believe that our initial assumption is not unreasonable.

Finally, we assume that the zonally asymmetric pattern is stationary in time. @silvestri2009 suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979-2018) so it's unlikely that we could observe significant changes. Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change (Figure A11).

@@

We perform linear regression to quantify the association between the SAM indices and other variables. Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions. To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Asymmetric SAM ($SAM_a$) and Symmetric SAM ($SAM_s$) we fit the ecuation

$$
X(\lambda, \phi, t) = \alpha(\lambda, \phi) SAM_a + \beta(\lambda, \phi) SAM_s + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
$$

where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear coefficients, $X_0$ and $\epsilon$ are the constant and error terms. From this equarion, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; in other words, it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and viceversa for $\beta$.

At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions. In such case, using naive p-values to test for significance leads to misleading results [@walker1914;@katz1991]. While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue. Therefore, p-values showed in regression fields are all adjusted following @benjamini1995.

When performing a separate regression for each trimester (DJF, MAM, JJA, SON) we first average the relevant variables to obtain a single value for each year and each trimester.

@@

```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM and density estimates.", fig.width=6, fig.height=3, fig.env="figure*"}

indices %>% 
  .[lev %in% c(50, 700) & term != "full"] %>% 
  ggplot(aes(estimate_norm)) +
  geom_density(aes(color = term)) +
  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam, guide = "none") +
  scale_y_continuous(NULL, trans = scales::reverse_trans(), labels = NULL, breaks = NULL) +
  scale_x_continuous(NULL, labels = NULL, breaks = NULL) +
  coord_flip()  +
  theme(strip.text = element_blank()) +

  indices %>% 
  .[lev %in% c(50, 700) & term != "full"] %>% 
  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
  ggplot(aes(time, estimate_norm)) +
  geom_line(aes(color = term)) +
  # ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
  #                           aes(color = term, width = ..density.., x = min(time - 1e7))) +
  # geom_smooth(aes(color = term), method = "loess", n = 481, span = 10*12/481, method.args = list(degree = 1)) +
  # geom_point(data = ~.x[lev == 50][time == as.Date("1987-11-01") & term == "sym" | 
  #                         time == as.Date("1987-08-01") & term == "asym"]) +
  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
  scale_y_continuous(NULL) +
  # scale_x_datetime(NULL, expand = c(0, 0)) +
  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
  tag_facets() +
  plot_layout(widths = c(1, 4))
```

The temporal evolution of the Assymmetric and Symmetric SAM was firstly asssesed. Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates. We selected these two levels as representative of the tropospheric and stratospheric variability respectively. As will be shown later, both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.

Month-to-month variability is evident for both indices, with noisy variations in the low frequency. At first glance the series can be distinguished by their distributions. Compared to the stratospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound. The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.

The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A3). This might suggests a link between stratospheric SAM variability and QBO. There is a local peak at 2 years in the periodigram of the tropospheric Symmetric SAM also, although it's not statistically significant. In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.

@@

From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated. Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987). We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care). At the same time, one-month-lag correlations increase with height. As a consequence, statospheric Asymmetric index tend to precede corresponding Symmetric index.

@@

Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices. For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposfere. Above 100 hPa correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability. Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere. Examining panels b and c, we see that the Asymemtric and Symmetric SAM  share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour. Stratospheric coherency is stronger for the Asymmetric SAM than the Symemtric SAM. The stratospheric Symmetric SAM seems to connect more strongly to the trosposphere than the Asymmetric SAM.

@@

```{r trends, fig.cap = "Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
annual_trend <- indices %>%
  copy() %>% 
  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  .[, season := "Annual"]

indices %>% 
  copy() %>% 
  # .[time > as.Date("1990-01-01")] %>%
  .[, season := season(time)] %>% 
  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  rbind(annual_trend) %>% 
  .[, estimate := estimate*3600*24*365*10] %>% 
  .[, std.error := std.error*3600*24*365*10] %>% 
  setnames("type", "term") %>% 
  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
  ggplot(aes(lev, estimate)) +
  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
              fill = "#95a3ab", color = "black", 
              alpha = .6, size = 0.1) +
  geom_line() +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  geom_hline(yintercept = 0) +
  scale_x_level() +
  scale_y_continuous("Standard deviations per decade") +
  coord_flip() +
  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
  tag_facets("cr") 
```

The trends for each of the indices (Full, Symmetric, Assymetric) were evaluate for the whole period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters. The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throught the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa. The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in in Summer, where the 100 hPa maximum is much more defined. Positive trends have been documented by previous studies [e.g. @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.

By separating the SAM signal in its Asymmetric and Symmetric parts, we can not only see that these trends are almost entirely due to the Symmetric component (colum b vs. column c), but in some cases the trends become more clear. In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`). In Autumn, using the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index.

@@

```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef. Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
indices_2d_yearly %>% 
  .[lat <= -20] %>% 
  .[lev %in% c(50, 700)] %>%
  .[, env := metR::WaveEnvelope(estimate), by = .(lat, term, lev)] %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
  geom_contour_tanaka(aes(z = estimate),
                      breaks = AnchorBreaks(0, exclude =  0),
                      size = 0.3,
                      global.breaks = FALSE) +
  # geom_contour2(aes(z = env)) +
  geom_qmap(~.x[lat <= -20]) +
  geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
                               lat = -49, term = "asym", lev = 700), 
             shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
  scale_x_longitude() +
  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
                       limits = c(-45, 45),
                       oob = scales::squish,
                       breaks = AnchorBreaks(0, 10, 0)) +
  scale_linetype(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
  tag_facets("cr") +
  axis_labs_smol  +
  no_grid
```

@@

In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3. The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures. The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel c.2). The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure. Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.

@@

The amplitude of first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean. Column `r get_col(term == "asym")` shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel `r get_col(term== "asym" & lev == 50)`), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel `r get_col(term== "asym" & lev == 700)`). Looking at panel `r get_tag(term == "asym" & lev == 700)` from Figure \ref{fig:2d-regr}, it becomes apparent that zonal wavess 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Easten Hemisphere.

@@

To analyse the vertical structure of the geopotential anomalies asociated with the asymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}). The geopotential anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM. The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes. Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.

@@

The structures shown in panels a and b in  Figure \ref{fig:vertical-regression} are surprisignly robust to the choice of index level. For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a. Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b. The patterns mainly change in amplitude.

```{r enso-cor}
oni <- rsoi::download_oni(TRUE, data_path("derived", "oni.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), value = ONI)]

mei <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), value = MEI)]

soi <- rsoi::download_soi(TRUE, data_path("derived", "soi.csv")) %>% 
  as.data.table() %>% 
  .[, .(time = lubridate::as_datetime(Date), value = -SOI)]

enso <- rbindlist(list(oni = oni, soi = soi, mei = mei), idcol = "index")

enso_sam_cor <- enso %>% 
  .[indices, on = "time", allow.cartesian = TRUE] %>% 
  .[, with(cor.test(estimate, value), 
           list(cor = estimate, 
                low = conf.int[1], 
                upp = conf.int[2])), by = .(lev, term, index)]

# ggplot(enso_sam_cor, aes(lev, cor, group = interaction(term, index))) +
#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
#   geom_line(aes(color = term)) +
#   scale_x_level() +
#   coord_flip()

cor_texts <- enso_sam_cor[lev == 700, 
                         paste0(signif(cor, 2)), # " (CI: ", signif(low, 2), "–", signif(upp, 2), ")"), 
                         by = .(term, index)] %>% 
  dcast(index~term)

enso_sam_partial <- enso %>% 
  .[indices_wide[lev == 700], on = "time"] %>% 
  .[, partial_cor(value, asym, sym), by = index] %>% 
  .[, partial_correlation := signif(partial_correlation, 2)] %>% 
  dcast(index~term)
```
